Metabolomics in CF and Healthy Controls

### 2022-02-08 Christine Chan/CF Metabolomics/chan_cfrd_metabolomics.Rmd
Update chan_cfrd_metabolomics.Rmd
moderated t tests

---
title: ""
author: "Tim Vigers & Laura Pyle"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
library(Hmisc)
library(mixOmics)
library(combinat)
library(tidyverse)
library(parallel)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD//PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
}
knitr::opts_knit$set(root.dir = home_dir)
```

@@

# Import metabolomics
metab_pos = read.csv("./Christine Chan/CF Metabolomics/Data_Raw/Christine Chan - CF-Diabetes - C18 Pos.csv",na.strings = "")
metab_neg = read.csv("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Christine Chan/CF Metabolomics/Data_Raw/Christine Chan - CF-Diabetes - C18 Neg.csv",na.strings = "")
hilic = read.csv("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Christine Chan/CF Metabolomics/Data_Raw/Christine Chan - CF-Diabetes - HILIC Pos.csv",na.strings = "")
# Clean
## Positive
metab_pos = clean_metab(metab_pos,prefix = "P")
compounds = metab_pos[[2]]
## Negative
metab_neg = clean_metab(metab_neg,prefix = "N")
compounds = rbind(compounds,metab_neg[[2]])
## HILIC
hilic = clean_metab(hilic,prefix = "H")
compounds = rbind(compounds,hilic[[2]])
# Merge
df = df %>% filter(subject_id %in% metab_pos[[1]]$subject_id) %>%
  left_join(.,metab_pos[[1]],by = "subject_id") %>% 
  left_join(.,metab_neg[[1]],by = "subject_id") %>%
  left_join(.,hilic[[1]],by = "subject_id")
# Larger groups
df$combined = df$group_category.factor
levels(df$combined) = c("HC & CFNGT","HC & CFNGT","CFAGT & CFRD", "CFAGT & CFRD")
# Remove low variance
df = df[,-nearZeroVar(df)$Position]
# Matrices for mixOmics
X = df %>% select(P1:H651)
Y_all = df$group_category.factor
Y_two = df$combined
```

@@

Permutation p value for component 1 is `r round((1 + sum(aucs$comp1 > real_auc_1))/perms,3)`

@@

Permutation p value for component 1 is `r round((1 + sum(aucs$comp2 > real_auc_2))/perms,3)`

>>>>

---
title: "Metabolomics in CF and Healthy Controls"
author: "Tim Vigers & Laura Pyle"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
---

```{r setup, include=FALSE}
library(Hmisc)
library(mixOmics)
library(combinat)
library(tidyverse)
library(parallel)
library(limma)
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
if (Sys.info()["sysname"] == "Linux"){
  home_dir = "~/UCD//PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
} else if (Sys.info()["sysname"] == "Darwin"){
  home_dir = "/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF"
}
knitr::opts_knit$set(root.dir = home_dir)
```

@@

# Import metabolomics
metab_pos = read.csv("./Christine Chan/CF Metabolomics/Data_Raw/Christine Chan - CF-Diabetes - C18 Pos.csv",na.strings = "")
metab_neg = read.csv("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Christine Chan/CF Metabolomics/Data_Raw/Christine Chan - CF-Diabetes - C18 Neg.csv",na.strings = "")
hilic = read.csv("/Volumes/PEDS/RI Biostatistics Core/Shared/Shared Projects/Vigers/CF/Christine Chan/CF Metabolomics/Data_Raw/Christine Chan - CF-Diabetes - HILIC Pos.csv",na.strings = "")
# Clean
## Positive
metab_pos = clean_metab(metab_pos,prefix = "P")
compounds = metab_pos[[2]]
## Negative
metab_neg = clean_metab(metab_neg,prefix = "N")
compounds = rbind(compounds,metab_neg[[2]])
## HILIC
hilic = clean_metab(hilic,prefix = "H")
compounds = rbind(compounds,hilic[[2]])
# Merge
df = df %>% filter(subject_id %in% metab_pos[[1]]$subject_id) %>%
  left_join(.,metab_pos[[1]],by = "subject_id") %>% 
  left_join(.,metab_neg[[1]],by = "subject_id") %>%
  left_join(.,hilic[[1]],by = "subject_id")
# Larger groups
df$combined = df$group_category.factor
levels(df$combined) = c("HC & CFNGT","HC & CFNGT","CFAGT & CFRD", "CFAGT & CFRD")
levels(df$group_category.factor) = c("HC","CFNGT","CFAGT","CFRD")
# Remove low variance
df = df[,-nearZeroVar(df)$Position]
# Matrices for mixOmics
X = df %>% select(P1:H651)
Y_all = df$group_category.factor
Y_two = df$combined
```

@@

Permutation p value for component 1 is `r round(sum(aucs$comp1 > real_auc_1)/perms,3)`

@@

Permutation p value for component 1 is `r round(sum(aucs$comp2 >= real_auc_2)/perms,3)`

# Moderated t tests

```{r warning=FALSE,message=FALSE}
l = t(log(X))
design = model.matrix(~group_category.factor,df)
fit = lmFit(l,design)
fit <- eBayes(fit)
res = topTable(fit,number = nrow(l))
colnames(res)[1:3] = c("CFNGT","CFAGT","CFRD")
res = res[res$adj.P.Val < 0.05 & !is.na(res$adj.P.Val),]
res$Compound = compounds$Compound[match(rownames(res),compounds$compound_id)]
res = res %>% select(Compound,everything())
kable(res,row.names = F)
```

`r nrow(res)` metabolites were different in at least one of the glycemia groups. Healthy controls are the reference group in the table above.

```{r warning=FALSE,message=FALSE}
design = model.matrix(~combined,df)
fit = lmFit(l,design)
fit <- eBayes(fit)
res = topTable(fit,number = nrow(l))
res = res[res$adj.P.Val < 0.05 & !is.na(res$adj.P.Val),]
res$Compound = compounds$Compound[match(rownames(res),compounds$compound_id)]
res = res %>% select(Compound,everything())
kable(res,row.names = F)
```

`r nrow(res)` metabolites were different between the combined glycemia groups. Healthy controls and CFNGT is the reference group in the table above, and logFC refers to the log fold change between groups.
